A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-10 latex/main.tex
Basic usage of 3SEQ data
Update

Update figures for new 3SEQ data

Update WIP

Sequencing errors in genome assemblies \cite{Turakhia2020},
read misalignments \cite{Hunt2024}, and
multi-nucleotide substitutions \cite{DeMaio2020issues}
can cause short regions of a sample genome to match more closely
to an alternative genome rather than the true parent.
Such genomic regions can be mistaken for evidence of recombination,
because the LS HMM counts each base separately.
We developed a quality control (QC) measure
which counts the number of ``loci'' (clusters of closely spaced sites)
supporting the left and right parents of a recombinant (STAR Methods),
and effectively identifies a range of issues.
Figure~\ref{fig:recombinant_qc} shows
the net number of supporting loci for the left and right parents
for all 855 recombination events.
The 501 recombinants having fewer than 4 net loci supporting each parent
are highly enriched for samples sequenced using the AmpliSeq V1 protocol
% https://github.com/jeromekelleher/sc2ts-paper/issues/808
(70\% of the recombinants despite making up 2\% of the Viridian dataset), and
71 (14\%) are associated with 6-base deletion in the Delta lineage,
known to be affected by amplicon dropout\cite{Sanderson2021}.
The 354 remaining recombination events that pass this QC filter
contain all of the recombinants associated with Pango X lineages
(Table~\ref{tab:pango_x_lineages}) and
the Jackson et al. recombinants (Document~\ref{sec:jackson_recombs}).

\subsection*{Quantification of evidence for recombination}
The LS HMM at the heart of sc2ts provides a simple and powerful means 
of quantifying the plausibility of a given recombination event.
To do this, we force the HMM to find a Viterbi solution for the 
putative recombinant sequence with no recombination by specifying
a large recombination penalty $k$ (Figure~\ref{fig:methodology}).
The difference between the numbers of mutations between
the two solutions---the number of mutations
``averted'' by recombination---is a measure of
the evidence in favor of recombination  (similar to 
idea of ``parsimony improvement'' in RIPPLES.)
Table~\ref{tab:pango_x_lineages} shows the number of averted mutations
for each of the 21 recombination events associated with Pango X lineages,
with support ranging from very strong in the case of XC
(the most parsimonious non-recombinant placement would
require 37 additional mutations) to marginal for XG and XW,
which have the minimal possible number of averted mutations.

@@

See Figure~\ref{fig:recombinant_evidence_rebar} for rebar,
Figure~\ref{fig:recombinant_evidence_ripples_p4} for RIPPLES p4
and Figure~\ref{fig:recombinant_evidence_ripples_p4} for RIPPLES p3.

@@

\subsection*{Recombination breakpoint intervals}

Although in practice
the exact location of a recombination breakpoint
is impossible to pinpoint,
bounds can be placed by
examining the parental and recombinant sequences.
In methods \cite{Zhou2023virusrecom,Alfonsi2024,Li2024CovRecomb},
and approaches (e.g., Pango designation),
based on lineage-associated mutations,
the breakpoints are limited in precision to these mutations.
Similarly, methods based on probabilistic representations of
mutational profiles (e.g.\cite{Varabyou2021rapid})
suffer from this limitation.
However, genealogical methods, such as sc2ts and RIPPLES,
based on identifying exact parental \emph{sequences},
are in principle maximally precise,
assuming that the parental sequences are well represented in the ARG.

We explored the genomic distribution of the breakpoint intervals of
the QC-passing recombination events in the ARG.
We found an overall preponderance of breakpoint intervals
towards the 3’ end of the genome,
particularly at the left boundary of the Spike gene
(Figure~\ref{fig:recombinant_intervals}A,B),
consistent with the patterns seen in previous
studies \cite{Bobay2020recombination,Turakhia2022,
Smith2023tracking,Li2024CovRecomb}.
While this might result from a higher recombination rate in this region,
it may also be due to the higher levels of polymorphism in the Spike gene,
allowing recombination endpoints to be estimated more precisely.

Using the size of a breakpoint interval,
we measured how precisely breakpoints could
be localized along the genome.
Across the QC-passing recombination events,
we found the median breakpoint interval size to be 2,207 bases
(Figure~\ref{fig:recombinant_intervals}C),
indicating that the breakpoints of half of the events
could be narrowed to a small region (less than $\sim$7\% of the genome),
while the uncertainty around the breakpoints of the other events
is considerably large.
Further, we found a negative relationship between
the size of a breakpoint interval and
the amount of divergence between recombinant parents,
which is measured by time to their MRCA
(Figure~\ref{fig:recombinant_intervals}D).
When the two parental lineages have diverged
for longer periods of time (the rightmost points),
more information becomes available
on either side of a breakpoint to pinpoint its location,
resulting in shorter intervals.
For instance, the breakpoint intervals tend to be the smallest
for events involving recombination between
the deeply divergent Omicron BA.1 and BA.2 sublineages
(the red points in Figure~\ref{fig:recombinant_intervals}D).

Another key feature of genealogical methods
is that they can also provide \emph{time} intervals
which reflect uncertainty
in the estimated timing of recombination events.
The time interval of a recombination node in the ARG is bounded by
the age of the youngest parent and
the age of the oldest child of the recombination node
(either given as collection dates or estimated using tsdate).
We found that for three quarters of the QC-passing events,
the time interval spans less than $\sim$2 months
(Figure~\ref{fig:recombinant_intervals}D),
with long time intervals suggesting that
a parental lineage or the recombinant lineage
might be unsampled for an extended period of time.
% Do we want to refer to the CIs around the node dates?
This result indicates that the ARG can pinpoint
when new recombinant lineages emerge with fine resolution.

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{recombinant_intervals}
\caption{
Recombination breakpoint intervals.
(A) Distribution of the breakpoint intervals of
354 recombination events along the genome
(annotated with gene labels on top).
Each line represents an interval colored by its size.
(B) Number of overlapping breakpoint intervals per site.
(C) Distribution of breakpoint interval size (bases).
(D) Relationship between breakpoint interval size and
time between the recombination event and the parental MRCA.
The black curve shows the theoretical expected relationship
between these quantities,
estimated by minimizing the sum of squared residuals
given a single parameter (the mutation rate) (STAR Methods).
(E) Distribution of time interval size (months).
}
\label{fig:recombinant_intervals}
\end{figure}

@@

\subsubsection*{Pandemic scale alignment storage}
To provide the efficient access to alignment data required for inference with sc2ts, we
developed a new format based on the VCF Zarr~\cite{Czech2025} specification.
Briefly, Zarr is a storage format for scientific data,
originally developed for the
\textit{Anopheles gambiae}
1000 Genomes Project~\cite{Anopheles2017genetic}, and
is now seeing widespread adoption across the 
sciences~\cite{Abernathey2021cloud,Moore2021ome,Gowan2022using,
Dhapola2022scarf,Virshup2023scverse,Baker2023emobject,Marconato2024spatialdata,Ruan2024image}.
Zarr is essentially a simple mechanism for storing n-dimensional array
data as a regular grid of compressed chunks, which provides 
high levels of data compression, efficient access across multiple
dimensions, and is ideally suited to modern cloud-based
deployments~\cite{Czech2025}.

@@

\subsubsection*{Sample filtering and alignment preprocessing}

@@

\subsubsection*{Calculation of Pango lineage distance using pangonet}
We define a ``pseudo-phylogenetic'' distance
to quantify the divergence between two Pango lineages.
The idea is to count the number of edges
separating two given Pango lineages in a pseudo-phylogeny
which relates all the designated Pango lineages.
To construct this pseudo-phylogeny, we used \textit{pangonet},
which mines the information on the Pango lineages
in the ``alias\_key.json'' and ``lineage\_notes.txt'' files
from the Pango designation issues
(\url{https://github.com/cov-lineages/pango-designation}).
If two Pango lineages are siblings in the pseudo-phylogeny,
the distance is defined as the sum of the number of edges from each lineage to their MRCA;
if they are non-siblings (i.e. one Pango is a descendant of the other),
the distance is defined as the number of edges separating them;
if they are identical, the distance is defined as zero.

@@

\subsubsection*{Estimating interval lengths}
Recombination intervals are imprecise
when the recombining parents are identical in sequence at sites spanning the breakpoint.
Here we estimate the interval lengths expected around a breakpoint,
given a genome-wide mutation rate $\mu$,
total time between two samples ($T$), and
sufficient support to detect the recombination event.
If a breakpoint occurs a fraction $P$ along a genome of length $L$ and
$k$ mutations have occurred to the left of the breakpoint,
then we expect the nearest mutation to occur $P L / (k + 1)$ bases to the left.
If we require $n$ mutations on both sides of the breakpoint,
we can calculate the expected value of the interval to the left
by summing across a Poisson-distributed number of mutations,
conditioned on having at least $n$ mutations:

A similar result holds for the right interval ($I_{right}$), but with $P$ replaced with $1–P$.
Summing the two intervals ($I_{left}+I_{right}$) gives the expected interval length.
As detectable recombination events are nearer the center of the chromosome,
we use $P=½$ in Figure X.

\subsubsection*{Analyzing global case counts}
To explore the timing of emergence of recombinants,
case count data were downloaded from
the COVID-19 Data Repository by the Center for Systems Science and Engineering (CSSE)
at Johns Hopkins University \cite{Dong2020} and binned by week from January, 2020 to March, 2023.
The proportion of each lineage $i$ in each week ($p_i$) was then inferred
from the Scorpio designation at each node within the ARG for that week.
Genetic variation was measured as the ``expected heterozygosity'' across lineages,
which is the chance that two lineages chosen at random
from that week had different Scorpio designations ($\sum_{i} \sum_{j>i} 2 p_i p_j$).
The linear regression models reported in \textbf{Figure} \ref{fig:recombinants_over_time} allowed for a non-zero intercept,
because there may be unreported cases even without any reported cases.
The slopes were highly significant, and
the p-values were confirmed by randomly permuting the predictor variable 10,000 times.
ANOVAs were conducted to determine which single predictor ($n$, $H$, $n \; H$, $n^2$, $n^2 H$) explained the most variation in recombination events by week.
These statistical tests and graphics were conducted
in \textit{Mathematica} 14.2 (Wolfram 2024) and available at GITHUB LINK.

@@

\subsection{Analysis of deletion events}
\label{sec:supplement_dels}
% [TODO This is too terse and too vague.
% Be concrete and precise.
%WHICH deletions are we talking about?]

% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
Although uncommon relative to single nucleotide mutations,
deletions are an important component of
SARS-CoV-2 evolution \cite{Jeronimo2023},
contributing to the emergence of VOCs by enhancing,
e.g., efficiency for cell entry \cite{Cantoni2022} and
potential for antibody escape \cite{McCarthy2021}.
Using the ARG, we identified major deletion events
(which are propagated to many descendants)
as well as recurrent deletions
(which occur on at least two branches) in the ARG.
In Table~\ref{tab:major_dels},
we list the major deletion events in the ARG
which are each inherited by at least 10,000 descendant samples; and
in Table~\ref{tab:rec_dels},
we list the recurrent deletions in the ARG
which are each observed in at least 10,000 (0.4\%) samples
% [FIXME define how we measured this].
(i.e., for a deletion caused by multiple events in the ARG,
the total of all the samples inheriting the deletion across those events).

@@

\subsection{Imputation comparison of sc2ts and UShER}
\label{sec:supplement_sc2ts_usher_imputation}
Both sc2ts and UShER support missing data in input alignments, and effectively
``hard call'' these values when the samples are integrated into the tree/ARG. 
In doing so, we impute the missing and ambiguous data for these samples.
We evaluate the imputation performance of sc2ts and UShER by comparing their
outputs on the Viridian alignments.

The total aligned dataset for 2,475,418 samples and 27,507 sites 
in the sc2ts and UShER intersection ARGs represents 63.41GiB of nucleotide calls. 
Of the 162,838,461 missing data calls (Ns) in the alignments,
sc2ts and UShER disagreed in their imputed values for only 203,579 (0.125\%).
Additionally, 956,859 calls made use of the IUPAC uncertainty codes. 
Of these sc2ts imputed 435 (0.05\%) incorrectly
(i.e., with a base that is not compatible with the ambiguity code).
UShER imputed only 7 calls (0.007\%) from this set incorrectly,
likely due to special handling of
ambiguous bases in the parsimony calculations.

@@

\subsection{Pango lineage ``origination'' events}
\label{sec:supplement_pango_lineages}
To analyze how well the ARG reflects the phylogenetic structure implicit in the
Pango lineage naming system, we considered the Pango assignments generated by
Pangolin on the alignments for each node in the ARG (STAR Methods). If the ARG
perfectly reflected the Pango lineage structure, each lineage would form a
clade descending from a single originating node with that label. We identified
putative origination nodes among those labeled with a given lineage as the node
with the maximum number of descendant samples and the earliest time (if there
is a tie).

@@

>>>>

Sequencing errors in genome assemblies \cite{Turakhia2020},
read misalignments \cite{Hunt2024}, and
multi-nucleotide substitutions \cite{DeMaio2020issues}
can cause short regions of a sample genome to match more closely
to an alternative genome rather than the true parent.
Such genomic regions can be mistaken for evidence of recombination,
because the LS HMM counts each base separately.
We developed a quality control (QC) measure
which counts the number of ``loci'' (clusters of closely spaced sites)
supporting the left and right parents of a recombinant (STAR Methods),
and effectively identifies a range of issues.
Figure~\ref{fig:recombinant_qc} shows
the net number of supporting loci for the left and right parents
for all 855 recombination events.
The 501 recombinants having fewer than 4 net loci supporting each parent
are highly enriched for samples sequenced using the AmpliSeq V1 protocol
% https://github.com/jeromekelleher/sc2ts-paper/issues/808
(70\% of the recombinants despite making up 2\% of the Viridian dataset), and
71 (14\%) are associated with 6-base deletion in the Delta lineage,
known to be affected by amplicon dropout\cite{Sanderson2021}.
The 354 remaining recombination events that pass this QC filter
contain all of the recombinants associated with Pango X lineages
(Table~\ref{tab:pango_x_lineages}) and
the Jackson et al. recombinants (Document~\ref{sec:jackson_recombs}).
This QC filter also correlates strongly with results of other 
recombination detection methods when run on the putative 
recombinant sequences (see next section).

\subsection*{Validating and evaluating plausibility of recombination events}
[THIS IS VERY MUCH WIP]

We validated the recombination events in the sc2ts ARG by running 
several recombination detection programs on the putative recombinant
sequences. [TODO generate supplemental table summarising results of 
valiation.]

We first ran 3SEQ~\cite{Lam2018improved} [etc etc].

To assess the evidence for these putative novel recombination events we 
ran CovRecomb~\cite{Li2024CovRecomb} and rebar on the corresponding sequences 
(STAR Methods). Of the 354 QC-passing sc2ts recombination events, 
90 are classified as 
recombinant by CovRecomb (all but 2 of the 510 QC-failing recombinants 
were classified as non-recombinant).
Rebar classified 203 of the QC-passing sc2ts recombination events, 
as recombinant, with only 10 of those failing QC being classified 
as recombinant.

The LS HMM at the heart of sc2ts provides a simple and powerful means 
of quantifying the plausibility of a given recombination event.
To do this, we force the HMM to find a Viterbi solution for the 
putative recombinant sequence with no recombination by specifying
a large recombination penalty $k$ (Figure~\ref{fig:methodology}).
The difference between the numbers of mutations between
the two solutions---the number of mutations
``averted'' by recombination---is a measure of
the evidence in favor of recombination  (similar to 
idea of ``parsimony improvement'' in RIPPLES.)
Table~\ref{tab:pango_x_lineages} shows the number of averted mutations
for each of the 21 recombination events associated with Pango X lineages,
with support ranging from very strong in the case of XC
(the most parsimonious non-recombinant placement would
require 37 additional mutations) to marginal for XG and XW,
which have the minimal possible number of averted mutations.

@@

See Figure~\ref{fig:recombinant_evidence_cov_recomb} CovRecomb
See Figure~\ref{fig:recombinant_evidence_rebar} for rebar,
Figure~\ref{fig:recombinant_evidence_ripples_p4} for RIPPLES p4
and Figure~\ref{fig:recombinant_evidence_ripples_p4} for RIPPLES p3.

@@

\subsection*{Precision of recombination event times and positions}

The location of a recombination breakpoint is generally
impossible to pinpoint exactly, and instead an interval
containing the breakpoint can be inferred by
examining the parental and recombinant sequences.
Some methods \cite{Zhou2023virusrecom,Alfonsi2024,Li2024CovRecomb}
and approaches (e.g., Pango designation) infer this
as the genomic distance between lineage-associated
mutations that differentiate the two parental \emph{lineages}.
Similarly, methods based on probabilistic representations of
mutational profiles \cite{Varabyou2021rapid}
can only be as precise as examination of the two
parent lineages allows.
However, genealogical methods (such as sc2ts and RIPPLES) are
based on identifying exact parental \emph{sequences},
so are in principle maximally precise.
The breakpoint interval of a recombinant is bounded by
two positions: a left position where the recombinant
matches the left parent sequence but not the right, and
a right position where the recombinant matches the right
parent sequence but not the left. 
How well the time and position of a recombination event
can be inferred thus depends on the divergence between the
parent sequences (measured as time to their MRCA).

The distribution of breakpoint intervals of
the QC-passing recombination events
shows an overall enrichment of breakpoints
towards the 3’ end of the genome,
particularly at the left boundary of the Spike gene
(Figure~\ref{fig:recombinant_intervals}A,B),
consistent with the patterns seen in previous
studies \cite{Bobay2020recombination,Turakhia2022,
Smith2023tracking,Li2024CovRecomb}.
While this might result from a higher recombination rate in this region,
it may also be due to the higher levels of polymorphism in the Spike gene,
making recombination events more likely to be detectable, and 
allowing breakpoint intervals to be estimated more precisely.
We found the median breakpoint interval size to be 2,207 bases
(Figure~\ref{fig:recombinant_intervals}C),
indicating that the breakpoints of half of the events
could be narrowed to a small region (less than $\sim$7\% of the genome),
while the uncertainty around the breakpoints of the other events
is considerably larger (up to $\sim$54\% of the genome).
As expected, we found a negative relationship between
the size of a breakpoint interval and
divergence between recombinant parents
(Figure~\ref{fig:recombinant_intervals}D),
which follows theoretical expectation (STAR Methods).
Breakpoint intervals tend to be the smallest
for events involving recombination between
the deeply divergent Omicron BA.1 and BA.2 sublineages
(the red points in Figure~\ref{fig:recombinant_intervals}D).

The time interval of a recombination node in the ARG is bounded by
the age of the youngest parent and
the age of the oldest child of the recombination node
(either given as collection dates or estimated using tsdate).
We found that for $\sim$75\% of the QC-passing recombination events,
the time interval size is less than $\sim$2 months
(Figure~\ref{fig:recombinant_intervals}D);
% Jackson samples in <= 7 day range:
% Group B ERR5058070 (5 day), singleton ERR5054123 (7 days)
for example, the time intervals of
the events associated with the Jackson et al.\ samples
span 5 to 48 days.
% Q3: 60 days, max: 259 days
Longer time intervals imply that a parental or a recombinant lineage
might have been unsampled for an extended period of time.

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{recombinant_intervals}
\caption{
Recombination intervals.
(A) Distribution of the breakpoint intervals of
354 recombination events along the genome
(annotated with gene labels on top).
Each line represents a breakpoint interval colored by its size.
(B) Number of overlapping breakpoint intervals per site.
(C) Length distribution of breakpoint intervals.
(D) Relationship between breakpoint interval size and
time between the recombination event and the parental MRCA.
The black curve shows
the theoretical expected relationship between these quantities,
which is calculated assuming a nucleotide substitution rate of
% N.B. This is a pre-VOC estimate.
% This study (https://pmc.ncbi.nlm.nih.gov/articles/PMC8807201/#sup1)
% estimated background rate including VOCs.
% 0.00065 * 30000 / 12 = 1.625 per month
$\sim2$ x $10^{-6}$ per site per day
\cite{Duchene2020,Ghafari2022,Markov2023},
or $\sim$0.06 per day (STAR Methods).
(E) Length distribution of time intervals around recombination events.
}
\label{fig:recombinant_intervals}
\end{figure}

@@

\subsubsection*{Pandemic-scale alignment storage}
To provide the efficient access to alignment data required for inference with sc2ts,
we developed a new format based on the VCF Zarr~\cite{Czech2025} specification.
Briefly, Zarr is a storage format for scientific data,
originally developed for the
\textit{Anopheles gambiae}
1000 Genomes Project~\cite{Anopheles2017genetic}, and
is now seeing widespread adoption across the 
sciences~\cite{Abernathey2021cloud,Moore2021ome,Gowan2022using,
Dhapola2022scarf,Virshup2023scverse,Baker2023emobject,
Marconato2024spatialdata,Ruan2024image}.
Zarr is essentially a simple mechanism for storing n-dimensional array
data as a regular grid of compressed chunks, which provides 
high levels of data compression, efficient access across multiple
dimensions, and is ideally suited to modern cloud-based
deployments~\cite{Czech2025}.

@@

\subsubsection*{Sample filtering and alignment pre-processing}

@@

\subsubsection*{Calculation of Pango lineage distance using pangonet}

We define a ``pseudo-phylogenetic'' distance
to quantify the divergence between two Pango lineages.
The idea is to count the number of edges
separating two given Pango lineages in a pseudo-phylogeny
which relates all the designated Pango lineages.
To construct this pseudo-phylogeny, we used \textit{pangonet},
which mines the information on the Pango lineages
in the ``alias\_key.json'' and ``lineage\_notes.txt'' files
from the Pango designation issues
(\url{https://github.com/cov-lineages/pango-designation}).
If two Pango lineages are siblings in the pseudo-phylogeny,
the distance is defined as
the sum of the number of edges from each lineage to their MRCA;
if they are non-siblings (i.e., one Pango is a descendant of the other),
the distance is defined as
the number of edges separating them;
if they are identical, the distance is defined as zero.

\subsubsection*{Expected lengths of recombination breakpoint intervals}

Recombination breakpoint intervals are imprecise
when the recombining parents are identical in sequence
at sites spanning the breakpoint.
Here, we estimate the interval lengths expected around a breakpoint,
given a genome-wide mutation rate $\mu$,
total time between two samples ($T$), and
sufficient support to detect the recombination event.
If a breakpoint occurs a fraction $P$ along a genome of length $L$ and
$k$ mutations have occurred to the left of the breakpoint,
then we expect the nearest mutation to occur $P L / (k + 1)$ bases to the left.
If we require $n$ mutations on both sides of the breakpoint,
we can calculate the expected value of the interval to the left
by summing across a Poisson-distributed number of mutations,
conditioned on having at least $n$ mutations:

A similar result holds for the right interval ($I_{right}$),
but with $P$ replaced with $1–P$.
Summing the two intervals ($I_{left}+I_{right}$)
gives the expected interval length.
As detectable recombination events
tend to be closer to the center of the genome,
we use $P=0.5$ in Figure~\ref{fig:recombinant_intervals}D.

@@

\subsubsection*{Analyzing global case counts}

To explore the timing of emergence of recombinants,
case count data were downloaded from
the COVID-19 Data Repository by the Center for Systems Science and Engineering (CSSE)
at Johns Hopkins University \cite{Dong2020} and binned by week from January, 2020 to March, 2023.
The proportion of each lineage $i$ in each week ($p_i$) was then inferred
from the Scorpio designation at each node within the ARG for that week.
Genetic variation was measured as the ``expected heterozygosity'' across lineages,
which is the chance that two lineages chosen at random
from that week had different Scorpio designations ($\sum_{i} \sum_{j>i} 2 p_i p_j$).
The linear regression models reported in \textbf{Figure} \ref{fig:recombinants_over_time} allowed for a non-zero intercept,
because there may be unreported cases even without any reported cases.
The slopes were highly significant, and
the p-values were confirmed by randomly permuting the predictor variable 10,000 times.
ANOVAs were conducted to determine which single predictor ($n$, $H$, $n \; H$, $n^2$, $n^2 H$) explained the most variation in recombination events by week.
These statistical tests and graphics were conducted
in \textit{Mathematica} 14.2 (Wolfram 2024) and available at GITHUB LINK.

@@

\subsection{Analysis of deletion events}
\label{sec:supplement_dels}

% See https://github.com/jeromekelleher/sc2ts-paper/issues/628
Although uncommon relative to single nucleotide mutations,
deletions are an important component of
SARS-CoV-2 evolution \cite{Jeronimo2023},
contributing to the emergence of VOCs by enhancing,
e.g., efficiency for cell entry \cite{Cantoni2022} and
potential for antibody escape \cite{McCarthy2021}.
Using the ARG, we identified major deletion events
(which are propagated to many descendants)
as well as recurrent deletions
(which occur on at least two branches) in the ARG.
In Table~\ref{tab:major_dels},
we list the major deletion events in the ARG
which are each inherited by at least 10,000 descendant samples; and
in Table~\ref{tab:rec_dels},
we list the recurrent deletions in the ARG
which are each observed in at least 10,000 (0.4\%) samples
% [FIXME define how we measured this].
(i.e., for a deletion resulting from multiple events in the ARG,
this is the total samples inheriting the deletion).

@@

\subsection{Imputation comparison of sc2ts and UShER}
\label{sec:supplement_sc2ts_usher_imputation}

Both sc2ts and UShER support missing data in input alignments, and effectively
``hard call'' these values when the samples are integrated into the tree/ARG. 
In doing so, we impute the missing and ambiguous data for these samples.
We evaluate the imputation performance of sc2ts and UShER by comparing their
outputs on the Viridian alignments.

The total aligned dataset for 2,475,418 samples and 27,507 sites 
in the sc2ts and UShER intersection ARGs represents
63.41 GiB of nucleotide calls.
Of the 162,838,461 missing data calls (Ns) in the alignments,
sc2ts and UShER disagreed in their imputed values for only 203,579 (0.125\%).
Additionally, 956,859 calls made use of the IUPAC uncertainty codes. 
Of these sc2ts imputed 435 (0.05\%) incorrectly
(i.e., with a base that is not compatible with the ambiguity code).
UShER imputed only 7 calls (0.007\%) from this set incorrectly,
likely due to special handling of
ambiguous bases in the parsimony calculations.

@@

\subsection{Pango lineage ``origination'' events}
\label{sec:supplement_pango_lineages}

To analyze how well the ARG reflects the phylogenetic structure implicit in the
Pango lineage naming system, we considered the Pango assignments generated by
Pangolin on the alignments for each node in the ARG (STAR Methods). If the ARG
perfectly reflected the Pango lineage structure, each lineage would form a
clade descending from a single originating node with that label. We identified
putative origination nodes among those labeled with a given lineage as the node
with the maximum number of descendant samples and the earliest time (if there
is a tie).

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{recombinant_evidence_cov_recomb}
\caption{Things}
\label{fig:recombinant_evidence_cov_recomb}
\end{figure}
